15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
update introduction pt. 4

In combination, increased canopy tree mortality and reduced canopy species recruitment in the understory is likely to have profound impacts on deciduous forests of eastern North America.  When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a]. In the mid-Atlantic region of the eastern United States, forests face a severe regeneration debt; there are insufficient juveniles of canopy tree species to replace the mature cohort of that species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023]. The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., *Acer* spp., *Fagus grandifolia*) than historically dominated much of the region (e.g., *Quercus* spp., *Carya* spp., [@nowacki_climate_2015]; Thompson *et al.* 2013), but even these are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].

>>>>

Increased canopy tree mortality in combination with reduced canopy species recruitment in the understory is likely to have profound impacts on the deciduous forests of eastern North America.  When global change pressures (*e.g.*, climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, after which there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a]. In the mid-Atlantic region of the eastern US, forests face a severe regeneration debt: there are insufficient juveniles of canopy tree species to replace the mature cohort of that species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023]. The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., maple, *Acer* spp.; American beech, *Fagus grandifolia*) than historically dominated much of the region (e.g., oak, *Quercus* spp.; hickory, *Carya* spp.; [@nowacki_climate_2015; @thompson_four_2013], but even these species associated with mesic conditions are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].
